Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-08-03 submission/manuscript.tex
update abstract with Ada suggestions

Marc A Sze\({^1}\), Nicholas A Lesniak\({^1}\), Mack T Ruffin
IV\({^2}\), Patrick D. Schloss\({^1}\)\({^\dagger}\)

@@

\textbf{Background.} Colorectal cancer (CRC) is a continuing health
concern with the majority of the risk for developing disease being due
to environmental factors. The microbiota is one of these environmental
factors with certain bacterial community members being associated with
CRC and other taxa being associated to colons without tumors. Some of
the bacterial species in taxa associated to colons without tumors can
use fiber to produce short-chain fatty acids (SCFAs) that can inhibit
tumor growth in model systems. However, the data supporting the
importance of SCFAs in human CRC is less certain. Here, we test the
hypothesis that SCFA concentrations are different in individuals with
colorectal tumors.

\textbf{Methods.} We analyzed a cross-sectional (n=490) and longitudinal
pre- and post-treatment (n=67) group for their concentrations of
acetate, butyrate, and propionate. Analysis also included imputed gene
relative abundance with PICRUSt, metagenomic sequencing on a subset
(n=85) of the total cross-sectional group, and tumor classification and
SCFA prediction models using Random Forest.

\textbf{Results.} No difference in SCFA concentrations were found
between individuals without tumors and patients with adenomas or
carcinomas (P-value \textgreater{} 0.15). Using metagenomic sequencing,
there was also no difference in genes involved with SCFA synthesis
between individuals without tumors and patients with adenomas or
carcinomas (P-value \textgreater{} 0.70). Finally, there was no
difference between classification models with or without SCFAs in their
ability to predict patients with adenomas or carcinomas versus
individuals without tumors (P-value \textgreater{} 0.05).

\textbf{Conclusions.} Although our data does not support the hypothesis
that SCFAs are different in individuals that have colorectal tumors,
there may be context specific scenarios where SCFAs may still be
beneficial for treatment of CRC. Alternatively, our observations also
support the hypothesis that there may be other mechanisms that have not
been thoroughly investigated that are more involved with the development
of human CRC.

@@

\textbf{Total significant positive correlations between OTU relative
abundance and SCFA concentration were similar for individuals without
tumors and patients with adenomas or carcinomas.} Having found no
difference between individuals without tumors and patients with adenomas
or carcinomas in SCFA concentrations or genes coding for enzymes
involved with SCFA synthesis, we next investigated if specific OTUs
correlated with SCFA concentrations. Using Spearmans rho we found that
the majority of significant OTU correlations were to taxa from
\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
{[}Figure 3 \& Table S5{]}. A similar pattern was observed when using
high/low SCFA groups based on the overall median concentration for that
specific SCFA {[}Figure S1 \& Table S6{]}. There was a noticeablely
higher number of significant negative correlations associated with
patients with adenomas for all SCFAs tested {[}Figure 3{]}. In
particular, OTUs from the \emph{Ruminococcaceae} family had the largest
share of these negative correlations within patients with adenomas
{[}Figure 3{]}. Although patients with adenomas had more positive
correlations between OTUs and SCFA concentrations, their total number
was more similar to individuals without tumors or patients with
carcinomas versus the analogous comparison for the number of negative
correlations {[}Figure 3{]}. The number of positive correlations between
OTUs and SCFA concentrations was similar between individuals without
tumors and patients with a carcinoma {[}Figure 3{]}. Overall, these
results suggest that the resident taxa that may change the most due to
colon tumors may not be ones that are responsible for the production of
acetate, butyrate, or propionate.

@@

The observations from this study do not support the hypothesis that SCFA
concentrations are different in individuals with tumors. Whether we
directly measured the SCFA concentration or investigated genes
associated with their production, no difference could be identified
between individuals without tumors and patients with adenomas or
carcinomas {[}Figure 1 \& 2{]}. Although there were differences in the
number of significant correlations between SCFA concentration and OTU
relative abundance based on whether individuals did not have tumors, had
an adenoma, or had a carcinoma, SCFA concentrations did not provide
increased model accuracy for tumor classification {[}Figure 3-4 \&
S1{]}. Instead, SCFA concentrations provided similar information to what
specific OTUs were already providing to the tumor classification models
{[}Figure 4{]}. Additionally, when models using OTU relative abundance
to classify SCFA concentrations were assessed, the OTUs that classified
to \emph{Clostridiales}, \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} were not the same as the OTUs that classified to
these taxa in the tumor models {[}Figure 4-5{]}. Collectively, our
observations suggest that resident taxa from \emph{Clostridiales},
\emph{Lachnospiraceae}, and \emph{Ruminococcaceae}, that are different
between individuals without tumors and patients with adenomas or
carinomas, are not that same as those involved with SCFA production.

Although SCFAs have been shown to be anti-tumorigenic, most of these
studies have been performed in model systems (16, 17). Additionally,
many of the \emph{in vivo} studies investigate proxies such as fiber
supplementation rather than SCFAs directly (14). Although it is well
known that breakdown products from gut bacteria results in SCFA
production, fiber effects on tumorigenesis may be through other
mechanisms in these \emph{in vivo} model systems. In addition to using
fiber supplementation as a proxy for SCFA treatment, observations on the
benefit of SCFAs in preventing tumorigenesis has been mixed. In previous
case/control studies lower SCFA concentrations between patients with
carcinomas and those without carcinomas was observed (19). Yet, this is
in contrast to multiple randomized-controlled trials that have found no
difference between patients who do and do not get fiber supplementation
to try and prevent tumor recurrence (20, 23). This could be because
fiber has other actions in humans besides providing an energy source for
bacteria to breakdown to make SCFAs. It could also be due to the fact
that SCFA concentrations and responses to fiber can vary quite a bit
between healthy individuals (24). This information taken together with
our observations would suggest that either indviduals who do not respond
to fiber supplementation would need to acquire these bacteria to achieve
a benefit or that SCFAs provide little to no benefit as an
anti-tumorigenic compound in colorectal cancer.

@@

One possible technical limitation is that a fecal sample may not be an
ideal type of biospecimen and that the effect SCFAs have on
tumorigenesis is only detected in the colon. However, this is unlikely
to be a major confounder. First, most \emph{in vivo} studies as well as
human studies have used fecal material in their analysis (18, 19).
Second, previous studies that measure SCFA changes after fiber
supplementation use fecal material to track these responses with a great
deal of success (24). Although there are limitations with the current
research on SCFAs and colorectal tumors, technical limitations are less
likely to be cause of this. Additionally, as mentioned earlier, our
observations along with the randomized controlled trials on fiber
supplementation in tumor recurrence (20) provide evidence that these
specific metabolites may not be protective or used as a general
treatment option in colorectal cancer. Yet, taxa that are associated
with SCFA production are consistently higher in indivdiuals without
colon tumors than patients with carcinomas (10, 11, 27).

The potential protection against colorectal cancers may not be from
SCFAs, even though taxa associated with their production are higher in
individuals without tumors versus patients with carcinomas (10, 11, 27).
Indeed our data would support the contention that the taxa are similar
to those associated with SCFA production but that these specific
microbes or OTUs themselves are not associated with SCFAs. In
particular, our results showing that different OTUs from the same
taxonomic classification are in tumor and SCFA Random Forest models
supports this hypothesis. This leads to the possibilty that protection
may be through two other routes. First, there could be a different
pathway or other less extensively studied metabolites that provides the
necessary protection against tumorigenesis. Alternatively, protection
may not occur via a metabolite but instead through niche exclusion of
mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (6,
12, 13)). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection (28) with
chronic inflammation replacing the role of antibiotics. Even though we
did not find lower concentrations of SCFAs associated with colorectal
tumors, we think that there are many exciting new avenues to explore
because of these results.

@@

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
previously described (29, 31, 32). In brief, sequences were quality
filtered and contigs created from the paired end reads. Any sequences
with ambiguous base calls were discarded. Contigs were then checked for
matches to the V4 region of the 16S rRNA gene using the SILVA database
(33). Chimeras were identifed and removed using UCHIME and OTUs
clustered at 97\% similarity (34). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptClust algorithm (35).

@@

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (44). Generally, a Kruskal-Walis rank sum
test with a Dunn's post-hoc test was used to assess differences between
the groups used. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (45). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, the imputed gene counts of
important mediators of SCFA synthesis was tested. Additionally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. From here we analyzed the number of
significant positive and negative correlations between OTU relative
abundance and SCFA concentrations in individuals without tumors and
patients with adenomas or carcinomas using Spearman's rho. Next, we
assessed whether OTUs alone or OTUs and SCFAs were better able to
classify individuals with and without tumors using Random Forest models.
Finally, models to classify high or low SCFA concentration based on the
median of each SCFA or the actual concentration using 16S rRNA gene
sequencing data was created using the Random Forest algorithm. For all
Random Forest models, the assessment of the most important variables was
based on the top 10 feautres (OTUs or SCFAs) using the mean decrease in
accuracy.

>>>>

Marc A. Sze\({^1}\), Nicholas A. Lesniak\({^1}\), Mack T. Ruffin
IV\({^2}\), Patrick D. Schloss\({^1}\)\({^\dagger}\)

@@

\textbf{Background.} Colorectal cancer (CRC) is increasing in prevalence
in individuals under 50 and because of this will be a continuing health
concern for the foreseeable future. The majority of the risk for
developing CRC is attributable to environmental factors. One of these
environmental factors is the microbiota, with certain bacterial
community members being associated with CRC and other taxa being
associated to individuals without tumors. Some of the bacterial species
in taxa associated to individuals without tumors can use fiber to
produce short-chain fatty acids (SCFAs) that inhibit tumor growth in
model systems. However, the data supporting the importance of SCFAs in
human CRC is less certain. Here, we test the hypothesis that SCFA
concentrations and taxa associated with their production are different
in individuals with colorectal tumors.

\textbf{Methods.} We analyzed a cross-sectional (n=490) and longitudinal
pre- and post-treatment (n=67) group for their fecal concentrations of
acetate, butyrate, and propionate. Analysis also included imputed gene
relative abundance with PICRUSt, metagenomic sequencing on a subset
(n=85) of the total cross-sectional group, and tumor classification and
SCFA prediction models using Random Forest.

\textbf{Results.} No difference in SCFA concentrations were found
between individuals without tumors and patients with adenomas or
carcinomas (P-value \textgreater{} 0.15). Using metagenomic sequencing,
there was also no difference in genes involved with SCFA synthesis
between individuals without tumors and patients with adenomas or
carcinomas (P-value \textgreater{} 0.70). Finally, there was no
difference between the ability of Random Forest models to predict
patients with adenomas or carcinomas versus individuals without tumors
(P-value \textgreater{} 0.05).

\textbf{Conclusions.} Although our data does not support the hypothesis
that fecal SCFA concentrations in patients in the general CRC population
are different, there still may be specific types of colorectal tumors
where SCFAs may be beneficial for treatment of CRC. Alternatively, our
observations also support the hypothesis that there may be other
metabolites or mechanisms (e.g.~bacterial niche exclusion) that may be
more protective against tumorigenesis and have not been thoroughly
investigated in the context of human CRC.

@@

\textbf{Total significant positive correlations between OTU relative
abundance and SCFA concentration were similar for individuals without
tumors and patients with adenomas or carcinomas.} Having found no
difference between individuals without tumors and patients with adenomas
or carcinomas in SCFA concentrations or genes coding for enzymes
involved with SCFA synthesis, we next investigated if specific OTUs
correlated with SCFA concentrations. Using Spearmans rho we found that
the majority of significant OTU correlations were to taxa from
\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
{[}Figure 3 \& Table S5{]}. A similar pattern was observed when using
high/low SCFA groups based on the overall median concentration for that
specific SCFA {[}Figure S1 \& Table S6{]}. There was a noticeably higher
number of significant negative correlations associated with patients
with adenomas for all SCFAs tested {[}Figure 3{]}. In particular, OTUs
from the \emph{Ruminococcaceae} family had the largest share of these
negative correlations within patients with adenomas {[}Figure 3{]}.
Although patients with adenomas had more positive correlations between
OTUs and SCFA concentrations, their total number was more similar to
individuals without tumors or patients with carcinomas versus the
analogous comparison for the number of negative correlations {[}Figure
3{]}. The number of positive correlations between OTUs and SCFA
concentrations was similar between individuals without tumors and
patients with a carcinoma {[}Figure 3{]}. Overall, these results suggest
that the resident taxa that may change the most due to colon tumors may
not be ones that are responsible for the production of acetate,
butyrate, or propionate.

@@

The observations from this study do not support the hypothesis that SCFA
concentrations are different in individuals with tumors. Whether we
directly measured the SCFA concentration or investigated genes
associated with their production, no difference could be identified
between individuals without tumors and patients with adenomas or
carcinomas {[}Figure 1 \& 2{]}. Although there were differences in the
number of significant correlations between SCFA concentration and OTU
relative abundance based on whether individuals did not have tumors, had
an adenoma, or had a carcinoma, SCFA concentrations did not provide
increased model accuracy for tumor classification {[}Figure 3-4 \&
S1{]}. Instead, SCFA concentrations provided similar information to what
specific OTUs were already providing to the tumor classification models
{[}Figure 4{]}. Additionally, when models using OTU relative abundance
to classify SCFA concentrations were assessed, the OTUs that classified
to \emph{Clostridiales}, \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} were not the same as the OTUs that classified to
these taxa in the tumor models {[}Figure 4-5{]}. Collectively, our
observations suggest that resident taxa from \emph{Clostridiales},
\emph{Lachnospiraceae}, and \emph{Ruminococcaceae}, that are different
between individuals without tumors and patients with adenomas or
carcinomas, are not that same as those involved with SCFA production.

Although SCFAs have been shown to be anti-tumorigenic, most of these
studies have been performed in model systems (16, 17). Additionally,
many of the \emph{in vivo} studies investigate proxies such as fiber
supplementation rather than SCFAs directly (14). Although it is well
known that breakdown products from gut bacteria results in SCFA
production, fiber effects on tumorigenesis may be through other
mechanisms in these \emph{in vivo} model systems. In addition to using
fiber supplementation as a proxy for SCFA treatment, observations on the
benefit of SCFAs in preventing tumorigenesis has been mixed. In previous
case/control studies lower SCFA concentrations between patients with
carcinomas and those without carcinomas was observed (19). Yet, this is
in contrast to multiple randomized-controlled trials that have found no
difference between patients who do and do not get fiber supplementation
to try and prevent tumor recurrence (20, 23). This could be because
fiber has other actions in humans besides providing an energy source for
bacteria to breakdown to make SCFAs. It could also be due to the fact
that SCFA concentrations and responses to fiber can vary quite a bit
between healthy individuals (24). This information taken together with
our observations would suggest that either individuals who do not
respond to fiber supplementation would need to acquire these bacteria to
achieve a benefit or that SCFAs provide little to no benefit as an
anti-tumorigenic compound in colorectal cancer.

@@

One possible technical limitation is that a fecal sample may not be an
ideal type of bio-specimen and that the effect SCFAs have on
tumorigenesis is only detected in the colon. However, this is unlikely
to be a major confounder. First, most \emph{in vivo} studies as well as
human studies have used fecal material in their analysis (18, 19).
Second, previous studies that measure SCFA changes after fiber
supplementation use fecal material to track these responses with a great
deal of success (24). Although there are limitations with the current
research on SCFAs and colorectal tumors, technical limitations are less
likely to be cause of this. Additionally, as mentioned earlier, our
observations along with the randomized controlled trials on fiber
supplementation in tumor recurrence (20) provide evidence that these
specific metabolites may not be protective or used as a general
treatment option in colorectal cancer. Yet, taxa that are associated
with SCFA production are consistently higher in individuals without
colon tumors than patients with carcinomas (10, 11, 27).

The potential protection against colorectal cancers may not be from
SCFAs, even though taxa associated with their production are higher in
individuals without tumors versus patients with carcinomas (10, 11, 27).
Indeed our data would support the contention that the taxa are similar
to those associated with SCFA production but that these specific
microbes or OTUs themselves are not associated with SCFAs. In
particular, our results showing that different OTUs from the same
taxonomic classification are in tumor and SCFA Random Forest models
supports this hypothesis. This leads to the possibility that protection
may be through two other routes. First, there could be a different
pathway or other less extensively studied metabolites that provides the
necessary protection against tumorigenesis. Alternatively, protection
may not occur via a metabolite but instead through niche exclusion of
mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (6,
12, 13)). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection (28) with
chronic inflammation replacing the role of antibiotics. Even though we
did not find lower concentrations of SCFAs associated with colorectal
tumors, we think that there are many exciting new avenues to explore
because of these results.

@@

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
previously described (29, 31, 32). In brief, sequences were quality
filtered and contigs created from the paired end reads. Any sequences
with ambiguous base calls were discarded. Contigs were then checked for
matches to the V4 region of the 16S rRNA gene using the SILVA database
(33). Chimeras were identified and removed using UCHIME and OTUs
clustered at 97\% similarity (34). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity using the OptClust algorithm (35).

@@

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (44). Generally, a Kruskal-Walis rank sum
test with a Dunn's post-hoc test was used to assess differences between
the groups used. Where appropriate Benjamini-Hochberg was used to
correct for multiple comparisons (45). First, we assessed differences in
SCFA concentrations measured by HPLC between individuals with normal
colons and patients with tumors (adenoma or carcinoma). We then analyzed
whether SCFA concentrations changed in patients with an adenoma or
carcinoma pre- versus post-treatment. Next, the imputed gene counts of
important mediators of SCFA synthesis was tested. Additionally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. From here we analyzed the number of
significant positive and negative correlations between OTU relative
abundance and SCFA concentrations in individuals without tumors and
patients with adenomas or carcinomas using Spearman's rho. Next, we
assessed whether OTUs alone or OTUs and SCFAs were better able to
classify individuals with and without tumors using Random Forest models.
Finally, models to classify high or low SCFA concentration based on the
median of each SCFA or the actual concentration using 16S rRNA gene
sequencing data was created using the Random Forest algorithm. For all
Random Forest models, the assessment of the most important variables was
based on the top 10 features (OTUs or SCFAs) using the mean decrease in
accuracy.
